### 2020-08-11 submission/manuscript.tex
Incorporate Pat's edits

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single university colony or vendor. We treated mice from 6
different sources (2 University of Michigan colonies and 4 vendors) with
a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
day later and then measured \emph{C. difficile} colonization levels
through 9 days post-infection. The microbiota were profiled via 16S rRNA
gene sequencing to examine the variation across sources and alterations
due to clindamycin treatment and \emph{C. difficile} challenge. While
all sources of mice were colonized 1-day post-infection, variation
emerged from days 3-7 post-infection with animals from some sources
colonized with \emph{C. difficile} for longer and at higher levels. We
identified bacteria that varied in relative abundance across sources and
throughout the experiment. Some bacteria were consistently impacted by
clindamycin treatment in all sources of mice including
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and 1-day post-infection community composition data.
With these models, we identified 4 bacteria that varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human inter-individual variation and can help identify
bacterial drivers of phenotypic variation in the context of CDIs.

@@

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although perturbation to the gut microbiota is an established risk,
there is variation in who becomes asymptomatically colonized, develops
an infection, or has an infection with adverse outcomes. \emph{C.
difficile} infection (CDI) mouse models are widely used to answer a
variety of \emph{C. difficile} pathogenesis questions. However, the
inter-individual variation between mice from the same breeding facility
is less than what is observed in humans. Therefore, we challenged mice
from 6 different breeding colonies with \emph{C. difficile}. We found
that the starting microbial community structures and \emph{C. difficile}
persistence varied by the source of mice. Interestingly, a subset of the
bacteria that varied across sources were associated with how long
\emph{C. difficile} was able to colonize. By increasing the
inter-individual diversity of the starting communities, we were able to
better model human diversity. This provided a more nuanced perspective
of \emph{C. difficile} pathogenesis.

@@

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs) due to there effect on the intestinal microbiota, but
there is variation in who goes on to develop severe or recurrent CDIs
after exposure (1, 2). Additionally, asymptomatic colonization, where
\emph{C. difficile} is detectable, but symptoms are absent has been
documented in infants and adults (3, 4). The intestinal microbiota has
been implicated in asymptomatic colonization (5, 6), susceptibility to
CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
much inter-individual microbiota variation contributes to the range of
outcomes observed after \emph{C. difficile} exposure relative to other
risk factors.

@@

We sequenced the 16S rRNA gene from fecal samples collected from these
mice after they acclimated to the University of Michigan animal housing
environment. We first examined the alpha diversity across the 6 sources
of mice. There was a significant difference in the richness (i.e.~number
of observed operational taxonomic units (OTUs)), but not Shannon
diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
structure of mice (Fig. 1C). Interactions between the source and cage
effects explained most of the observed variation between fecal
communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
\textless{} 0.001, Fig. 1C and Table S3). Mice that are co-housed tend
to have similar gut microbiotas due to coprophagy (41). Since mice
within the same source were housed together, it is not surprising that
the cage effect also contributed to the observed microbiota variation.
There were some differences between the 2 experiments we conducted, as
the experiment and cage effects significantly explained the observed
community variation for the Schloss and Young lab mouse colonies (Fig.
S2A-B and Table S4). However, most of the vendors also clustered by
experiment (Fig. S2C-D, F), suggesting there was some community
variation between the 2 experiments within each source, particularly for
Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
at the community level, we next identified the bacteria that varied
between sources of mice. There were 268 OTUs with relative abundances
that were significantly different between the sources (Fig. 1D and Table
S5). Though we saw differences between experiments at the community
level, there were no OTUs that were significantly different between
experiments within Schloss, Young, and Envigo mice (Data not shown). By
using mice from six sources we were able to increase the number of
microbiota communities to treat with clindamycin and challenge with
\emph{C. difficile}.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but cleared the pathogen within 9
days when treated with clindamycin alone (21, 43). All mice were treated
with 10 mg/kg clindamycin via intraperitoneal injection and one day
later challenged with 10\textsuperscript{3} \emph{C. difficile} 630
spores (Fig. 2A). The day after infection, \emph{C. difficile} was
detectable in all mice at a similar level (median CFU range:
2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15), indicating
clindamycin rendered all mice susceptible regardless of source (Fig.
2B). However, between 3 and 7 days post-infection, we observed variation
in \emph{C. difficile} levels across sources of mice (all
\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S6). This
suggested the mouse source was associated with \emph{C. difficile}
clearance. While the colonization dynamics were similar between the two
experiments, the Schloss mice took longer to clear in the 1st experiment
and the Envigo mice took longer to clear in the 2nd experiment (Fig.
S2A-B). The change in the mice's weight significantly varied across
sources of mice with the most weight lost two days post-infection (Fig.
2C and Table S7). There was also one Jackson and one Envigo mouse that
died between 1- and 3-days post-infection during the second experiment.
Mice obtained from Jackson, Taconic, and Envigo tended to lose more
weight, have higher \emph{C. difficile} CFU levels and take longer to
clear the infection compared to the other sources of mice (although
there was variation between experiments with Schloss and Envigo mice).
This was particularly evident 7 days post-infection (Fig. 2B-C, Fig.
S2C-D), when 57\% of the mice were still colonized with \emph{C.
difficile} (Fig. S2E). By 9 days post-infection the majority of the mice
from all sources had cleared \emph{C. difficile} (Fig. 2C) with the
exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in mouse microbiotas that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by microbiota variation across the 6
sources of mice. As expected, clindamycin treatment decreased the
richness and Shannon diversity across all sources of mice (Fig. 3A-B).
Interestingly, significant differences in diversity metrics between
sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after
clindamycin treatment, with Charles River mice having higher richness
and Shannon diversity than most of the other sources (Fig 3A-B and
Tables S1-2). The clindamycin treatment decreased the variation in
community structures between sources of mice. Source and cage effects
explained almost all of the observed variation between communities
(combined R\textsuperscript{2} = 0.99, \emph{P} \textless{} 0.001; Fig.
3C and Table S3). However, there were only 18 OTUs (Fig. 3D and Table
S8) with relative abundances that significantly varied between sources.
Next we identified the bacteria that shifted after clindamycin
treatment, regardless of source by analyzing paired fecal samples from
mice that were collected at baseline and after clindamycin treatment. We
identified 153 OTUs that were altered after clindamycin treatment in
most mice (Fig. 3E and Table S9). When we compared the list of
significant clindamycin impacted bacteria with the bacteria that varied
between sources post-clindamycin, we found 4 OTUs
(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
between sources also shifted with clindamycin treatment. For example,
\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased after clindamycin
treatment (Fig. 3D). These findings demonstrate that clindamycin had a
consistent impact on the fecal bacterial communities of mice from all
sources and only a subset of the OTUs continued to vary between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse microbiotas and
were also able to clear \emph{C. difficile} faster than some of the
other sources. Microbiota diversity did not explain the observed
variation in \emph{C. difficile} colonization across sources considering
the Young and Schloss mice had the lowest diversity after clindamycin
treatment and were able to clear \emph{C. difficile} earlier than
Jackson, Taconic and Envigo mice. Source and the interactions between
source and cage effects continued to explain most of the observed
community variation (combined R\textsuperscript{2} = 0.88; \emph{P}
\textless{} 0.001; Fig. 4D and Table S3). One day after \emph{C.
difficile} challenge, there were 44 OTUs (Fig. 4D and Table S10) with
significantly different relative abundances across sources.

Throughout the experiment the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Mice fecal samples from the same
source of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice still had not recovered to
their baseline community structure (Fig. 4E). Distance from the baseline
community between sources did not explain the variation in \emph{C.
difficile} clearance since Schloss and Young mice cleared \emph{C.
difficile} faster, but their communities were a greater distance from
baseline 7 days post-infection. In summary, mouse bacterial communities
varied significantly between sources throughout the course of the
experiment and a consistent subset of bacterial taxa remained different
between sources regardless of clindamycin and \emph{C. difficile}
challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we next wanted to
determine which taxa were influencing the variation in \emph{C.
difficile} colonization at day 7 (Fig. 2B, Fig. S2C). We trained three
L2-regularized logistic regression models with input bacterial community
data from the baseline (day = -1), post-clindamycin (day = 0), and
post-infection (day = 1) timepoints of the experiment to predict
\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
models were better at predicting \emph{C. difficile} colonization status
on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
The model based on the post-clindamycin (day 0) community OTU data
performed significantly better than the other models with an area under
the receiving operator characteristic curve (AUROC) of 0.78
(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons,
Table S13). Thus, we were able to use bacterial relative abundance data
from the time of \emph{C. difficile} challenge to differentiate mice
that had cleared \emph{C. difficile} before day 7 from the mice still
colonized with \emph{C. difficile} at that timepoint. This result
suggests the bacterial community's response to clindamycin treatment had
the greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
colonization 7 days post-infection, while 3 OTUs
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
Next, we examined whether any of the top 20 ranked OTUs from the
baseline (day 0) model were also important in the other 2 classification
models based on baseline (day -1) and 1 day post-infection community
data. We identified 6 OTUs (Enterobacteriaceae, Ruminococcaceae,
Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae)
that were important to the day 0 model and either the baseline or 1 day
post-infection models (Table S14). Thus, a subset of bacterial OTUs were
important for determining \emph{C. difficile} colonization dynamics
across multiple timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we examined whether the top 20 ranked OTUs from each model overlapped
with OTUs that varied between sources (Fig. 1D, 3D, 4D and Tables S5,
S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table
S9). We found a subset of OTUs that were important to the baseline (day
-1), post-clindamycin (day 0), and 1 day post-infection models and
overlapped with bacteria that varied between sources, were altered by
clindamycin treatment, or both (Fig. S4). Combining the overall
comparison results for the 3 models identified 14 OTUs associated with
source, 21 OTUs associated with clindamycin treatment, and 6 OTUs
associated with both (Fig. 5B). Several OTUs (\emph{Bacteroides,
Enterococcus, Enterobacteriaceae, Porphyromonadaceae}) that overlapped
with our previous analyses appeared across at least 2 models, so we
examined how the relative abundances of these OTUs varied over the
course of the experiment (Fig. 6). Throughout the experiment, there was
at least 1 timepoint where relative abundances of these OTUs
significantly varied between sources (Table S15). Interestingly, there
were no OTUs that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection, suggesting multiple
bacteria influence \emph{C. difficile} colonization dynamics. Together,
these results suggest the initial bacterial communities and their
responses to clindamycin influence the clearance of \emph{C. difficile}.

@@

By running our CDI model with mice from 6 different sources, we were
able to identify bacterial taxa that were unique to sources throughout
the experiment as well as taxa that were universally impacted by
clindamycin. We trained L2 logistic regression models with baseline (day
-1), post-clindamycin treatment (day 0), and 1-day post-infection fecal
community data that could predict whether mice cleared \emph{C.
difficile} by 7 days post-infection better than random chance. We
identified \emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae} (Fig. 6) as candidate bacteria within these
communities that were influencing variation in \emph{C. difficile}
colonization dynamics since these bacteria were all important in the
logistic regression models and varied by source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrated
clindamycin was sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

Other studies have taken similar approaches by using mice from multiple
sources to identify bacteria that either promote colonization resistance
or increase susceptibility to enteric infections (22, 23, 26--30). For
example, in the context of \emph{Salmonella} infections,
\emph{Enterobacteriaceae} and segmented filamentous bacteria have
emerged as protective (22, 27). A previous study with \emph{C.
difficile} identified an endogenous protective \emph{C. difficile}
strain LEM1 that bloomed after antibiotic treatment in mice from Jackson
or Charles River Laboratories, but not Taconic that protected mice
against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
we obtained mice from the same vendors, we checked all mice for
endogenous \emph{C. difficile} by plating stool samples that were
collected after clindamycin treatment. However, we did not identify any
endogenous \emph{C. difficile} strains prior to challenge, suggesting
there were no endogenous protective strains in the mice we received and
other bacterial taxa mediated the variation in \emph{C. difficile}
colonization across sources.

>>>>

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single source. We treated mice from 6 sources (2 University of
Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by
\emph{C. difficile} challenge 1 day later and then measured \emph{C.
difficile} colonization levels through 9 days post-infection. The
microbiota were profiled via 16S rRNA gene sequencing to examine the
variation across sources and alterations due to clindamycin treatment
and \emph{C. difficile} challenge. While all mice were colonized 1-day
post-infection, variation emerged from days 3-7 post-infection with
animals from some sources colonized with \emph{C. difficile} for longer
and at higher levels. We identified bacteria that varied in relative
abundance across sources and throughout the experiment. Some bacteria
were consistently impacted by clindamycin treatment in all sources of
mice including \emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and 1-day post-infection community composition data.
With these models, we identified 4 bacteria that were predictive of
whether \emph{C. difficile} cleared. They varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human inter-individual variation and can help identify
bacterial drivers of phenotypic variation in the context of CDIs.

@@

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although perturbation to the gut microbiota is an established risk,
there is variation in who becomes asymptomatically colonized, develops
an infection, or has adverse infection outcomes. Mouse models of
\emph{C. difficile} infection (CDI) are widely used to answer a variety
of \emph{C. difficile} pathogenesis questions. However, the
inter-individual variation between mice from the same breeding facility
is less than what is observed in humans. Therefore, we challenged mice
from 6 different breeding colonies with \emph{C. difficile}. We found
that the starting microbial community structures and \emph{C. difficile}
persistence varied by the source of mice. Interestingly, a subset of the
bacteria that varied across sources were associated with how long
\emph{C. difficile} was able to colonize. By increasing the
inter-individual diversity of the starting communities, we were able to
better model human diversity. This provided a more nuanced perspective
of \emph{C. difficile} pathogenesis.

@@

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs) due to there effect on the intestinal microbiota, but
there is variation in who goes on to develop severe or recurrent CDIs
after exposure (1, 2). Additionally, asymptomatic colonization, where
\emph{C. difficile} is detectable, but symptoms are absent, has been
documented in infants and adults (3, 4). The intestinal microbiota has
been implicated in asymptomatic colonization (5, 6), susceptibility to
CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
much inter-individual microbiota variation contributes to the range of
outcomes observed after \emph{C. difficile} exposure relative to other
risk factors.

@@

We sequenced the 16S rRNA gene from fecal samples collected from these
mice after they acclimated to the University of Michigan animal housing
environment. We first examined the alpha diversity across the 6 sources
of mice. There was a significant difference in the richness (i.e.~number
of observed operational taxonomic units (OTUs)), but not Shannon
diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
structure of mice (Fig. 1C). The source of mice and the interactions
between the source and cage effects explained most of the observed
variation between fecal communities (PERMANOVA combined
R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001, Fig. 1C and
Table S3). Mice that are co-housed tend to have similar gut microbiotas
due to coprophagy (41). Since mice within the same source were housed
together, it was not surprising that the cage effect also contributed to
the observed community variation. There were some differences between
the 2 experiments we conducted, as the experiment and cage effects
significantly explained the observed community variation for the Schloss
and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of
the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
there was some community variation between the 2 experiments within each
source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
After finding differences at the community level, we next identified the
bacteria that varied between sources of mice. There were 268 OTUs with
relative abundances that were significantly different between the
sources (Fig. 1D and Table S5). Though we saw differences between
experiments at the community level, there were no OTUs that were
significantly different between experiments within Schloss, Young, and
Envigo mice (all \emph{P} \textgreater{} 0.05). By using mice from six
sources we were able to increase the variation in the starting
communities to evaluate in a clindamycin-based CDI model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but clear the pathogen within 9 days
when treated with clindamycin alone (21, 43). All mice were treated with
10 mg/kg clindamycin via intraperitoneal injection and one day later
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores
(Fig. 2A). The day after infection, \emph{C. difficile} was detectable
in all mice at a similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 2B). However, between 3 and
7 days post-infection, we observed variation in \emph{C. difficile}
levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
0.019; Fig. 2B and Table S6). This suggested the mouse source was
associated with \emph{C. difficile} clearance. While the colonization
dynamics were similar between the two experiments, the Schloss mice took
longer to clear in the 1st experiment compared to the 2nd and the Envigo
mice took longer to clear in the 2nd experiment compared to the 1st
(Fig. S2A-B). The change in the mice's weight significantly varied
across sources of mice with the most weight loss occurring two days
post-infection (Fig. 2C and Table S7). There was also one Jackson and
one Envigo mouse that died between 1- and 3-days post-infection during
the second experiment. Mice obtained from Jackson, Taconic, and Envigo
tended to lose more weight, have higher \emph{C. difficile} CFU levels
and take longer to clear the infection compared to the other sources of
mice (although there was variation between experiments with Schloss and
Envigo mice). This was particularly evident 7 days post-infection (Fig.
2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
the mice from all sources had cleared \emph{C. difficile} (Fig. 2B) with
the exception of 1 Taconic mouse from the first experiment and 2 Envigo
mice from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in fecal communities that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by community variation across the 6 sources
of mice. As expected, clindamycin treatment decreased the richness and
Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics between sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
treatment, with Charles River mice having higher richness and Shannon
diversity than most of the other sources (Fig 3A-B and Tables S1-2). The
clindamycin treatment decreased the variation in community structures
between sources of mice. The source of mice and the interactions between
source and cage effects explained almost all of the observed variation
between communities (combined R\textsuperscript{2} = 0.99, \emph{P}
\textless{} 0.001; Fig. 3C and Table S3). However, there were only 18
OTUs (Fig. 3D and Table S8) with relative abundances that significantly
varied between sources. Next, we identified the bacteria that shifted
after clindamycin treatment, regardless of source by analyzing paired
fecal samples from mice that were collected at baseline and after
clindamycin treatment. We identified 153 OTUs that were altered after
clindamycin treatment in most mice (Fig. 3E and Table S9). When we
compared the list of significant clindamycin impacted bacteria with the
bacteria that varied between sources post-clindamycin, we found 4 OTUs
(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
between sources also shifted with clindamycin treatment. For example,
\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased in relative
abundance after clindamycin treatment (Fig. 3D). These findings
demonstrate that clindamycin had a consistent impact on the fecal
bacterial communities of mice from all sources and only a subset of the
OTUs continued to vary between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse communities and
were also able to clear \emph{C. difficile} faster than the other
sources, diversity did not explain the observed variation in \emph{C.
difficile} colonization across sources. The Young and Schloss mice had
the lowest diversity 1 day post-infection and were able to clear
\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
source of mice and the interactions between source and cage effects
continued to explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
Table S3). One day after \emph{C. difficile} challenge, there were 44
OTUs (Fig. 4D and Table S10) with significantly different relative
abundances across sources.

Throughout the experiment, the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Mice fecal samples from the same
source of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice had not recovered to their
baseline community structure (Fig. 4E). The distance to the baseline
community did not explain the variation in \emph{C. difficile} clearance
as the Schloss and Young mice had mostly cleared \emph{C. difficile},
but their communities were a greater distance from baseline 7 days
post-infection compared to the Jackson and Taconic mice that were still
colonized. In summary, mouse bacterial communities varied significantly
between sources throughout the course of the experiment and a consistent
subset of bacteria remained different between sources regardless of
clindamycin and \emph{C. difficile} challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we determined which taxa
were influencing the variation in \emph{C. difficile} colonization at
day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
regression models with input bacterial community data from the baseline
(day = -1), post-clindamycin (day = 0), and post-infection (day = 1)
timepoints of the experiment to predict \emph{C. difficile} colonization
status on day 7 (Fig. S3A-B). All models were better at predicting
\emph{C. difficile} colonization status on day 7 than random chance (all
\emph{P} \textless{} 0.001, Table S12). The model based on the
post-clindamycin (AUROC = 0.78) community OTU data performed
significantly better than the baseline (AUROC = 0.72) or the
post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
\textless{} 0.001 for pairwise comparisons, Fig. S3C and Table S13).
Thus, we were able to use bacterial relative abundance data from the
time of \emph{C. difficile} challenge to differentiate mice that had
cleared \emph{C. difficile} before day 7 from the mice still colonized
with \emph{C. difficile} at that timepoint. This result suggests the
bacterial community's response to clindamycin treatment had the greatest
influence on subsequent \emph{C. difficile} colonization dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
colonization 7 days post-infection, while 3 OTUs
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
Next, we examined whether any of the top 20 ranked OTUs from the
post-clindamycin (day 0) model were also important in the other 2
classification models based on baseline (day -1) and 1 day
post-infection community data. We identified 6 OTUs
(\emph{Enterobacteriaceae, Ruminococcaceae, Lactobacillus, Bacteroides,
Porphyromonadaceae, Erysipelotrichaceae}) that were important to the
post-clindamycin model and either the baseline or 1 day post-infection
models (Table S14). Thus, a subset of bacterial OTUs were important for
determining \emph{C. difficile} colonization dynamics across multiple
timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we identified the OTUs from each model that varied between sources (Fig.
1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
important to the 3 models identified 14 OTUs associated with source, 21
OTUs associated with clindamycin treatment, and 6 OTUs associated with
both (Fig. 5B). Together, these results suggest that the initial
bacterial communities and their responses to clindamycin influence the
clearance of \emph{C. difficile}.

Several OTUs (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae}) that overlapped with our previous analyses appeared
across at least 2 models, so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Across the
9 days post-infection, there was at least 1 timepoint when the relative
abundances of these OTUs significantly varied between sources (Table
S15). Interestingly, there were no OTUs that emerged as consistently
enriched or depleted in mice that were colonized past 7 days
post-infection, suggesting that multiple bacteria influence \emph{C.
difficile} colonization dynamics.

@@

Applying our CDI model with mice from 6 different sources, allowed us to
identify bacterial taxa that were unique to different sources as well as
taxa that were universally impacted by clindamycin. We trained L2
logistic regression models with baseline (day -1), post-clindamycin
treatment (day 0), and 1-day post-infection fecal community data that
could predict whether mice cleared \emph{C. difficile} by 7 days
post-infection better than random chance. We identified
\emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
(Fig. 6) as candidate bacteria within these communities that were
influencing variation in \emph{C. difficile} colonization dynamics since
these bacteria were all important in the logistic regression models and
varied by source, were impacted by clindamycin treatment, or both.
Overall, our results demonstrated clindamycin was sufficient to render
mice from multiple sources susceptible to CDI and only a subset of the
inter-individual microbiota variation across mice from different sources
was associated with the time needed to clear \emph{C. difficile}.

Other studies have taken similar approaches by using mice from multiple
sources to identify bacteria that either promote colonization resistance
or increase susceptibility to enteric infections (22, 23, 26--30). For
example, against \emph{Salmonella} infections, \emph{Enterobacteriaceae}
and segmented filamentous bacteria have emerged as protective (22, 27).
We found \emph{Enterobacteriaceae} increased in all sources of mice
after clindamycin treatment, facilitating \emph{C. difficile}
colonization. However, there was also variation in
\emph{Enterobacteriaceae} relative abundance levels between sources that
was associated with the variation in \emph{C. difficile} colonization
dynamics across sources. Thus, bacteria may have differential roles in
determining susceptibility depending on the type of bacterial infection.

Differences in CDI mouse model studies have been attributed to
intestinal microbiota variation across sources. For example, researchers
using the same clindamycin treatment and C57BL/6 mice had different
\emph{C. difficile} outcomes, one having sustained colonization (32),
while the other had transient (18), despite both using \emph{C.
difficile} VPI 10643. Baseline differences in the microbiota composition
have been hypothesized to partially explain the differences in
colonization outcomes and overall susceptibility to \emph{C. difficile}
after treatment with the same antibiotic (13, 31). A previous study with
\emph{C. difficile} identified an endogenous protective \emph{C.
difficile} strain LEM1 that bloomed after antibiotic treatment in mice
from Jackson or Charles River Laboratories, but not Taconic that
protected mice against the more toxigenic \emph{C. difficile} VPI10463
(26). Given that we obtained mice from the same vendors, we checked all
mice for endogenous \emph{C. difficile} by plating stool samples that
were collected after clindamycin treatment. However, we did not identify
any endogenous \emph{C. difficile} strains prior to challenge,
suggesting there were no endogenous protective strains in the mice we
received and other bacteria mediated the variation in \emph{C.
difficile} colonization across sources. The \emph{C. difficile} strain
used could also be contributing to the variation in \emph{C. difficile}
outcomes seen across different research groups. For example, a group
found differential colonization outcomes after clindamycin treatment,
with \emph{C. difficile} 630 and M68 infections eventually becoming
undetectable while strain BI-7 remained detectable up to 70 days
post-treatment (44).

The bacterial perturbations induced by clindamycin treatment have been
well characterized and our findings agree with previous CDI mouse model
work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
were associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 43--46). While we have demonstrated that
susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments.
